---
title: "Block Maxima Annual Return Level Exceedances"
author: "Samuel McDonnell"
date: "2024-04-25"
output: html_document
---

```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(tmap)
library(tidyr)

data <- readRDS("gtsr_daily_max_1979_2014.rds")
```

## Dunlaoghaire (DL) Example 

Look at Dun Laoghaire first as an example. 




From below the 5yr return level for DL is 0.937447. This can be used as a cross check for the code when running for all stations. 

```{r}
library(extRemes)
dun_laoghaire <- data %>% filter(station_name == "795") #795 dun laoghaire
dun_laoghaire_annual_max <- dun_laoghaire %>% group_by(year) %>% summarise(annual_max = max(surge_daily_max))

# Fit GEV distribution to block maxima
gev_block_fit <- fevd(dun_laoghaire_annual_max$annual_max, method = "MLE")

print("Block Maxima Method (GEV/GP):")
print(summary(gev_block_fit))

plot(gev_block_fit) ## plot block max fit (looks better, fits inside 

return.level(gev_block_fit, return.period = c(2,5,20, 100), do.ci=TRUE)
```
## Extremes analysis for all stations

```{r, messages = FALSE}
#BLOCK MAXIMA APPROACH

# Define function to estimate return levels for a given station
estimate_return_levels_block<- function(station_data_block) {
  # Fit GEV distribution to block maxima
  gev_block_fit <- fevd(station_data_block$surge_annual_max, method = "MLE")
  # Estimate return levels
  return_levels_block <- return.level(gev_block_fit, return.period = c(2,5,20, 100))
  
  # Return the return levels
  return(return_levels_block)
}

# Create an empty dataframe to store return level estimates for all stations
return_levels_all_block <- data.frame(station_name = character(),
                                 return_2yr = numeric(),
                                 return_5yr = numeric(),
                                 return_20yr = numeric(), 
                                 return_100yr = numeric(),
                                 stringsAsFactors = FALSE)

# Loop through each station in the subset (i = 16 for DunL)
for (i in 1:nlevels(data$station_name)) {
  # Subset data for the current station
  station_data_block <- data %>% filter(station_name == levels(data$station_name)[i]) %>% group_by(station_name, year) %>% summarise(surge_annual_max = max(surge_daily_max)) %>% ungroup() 
  
  # Estimate return levels for the current station
  return_levels_block <- estimate_return_levels_block(station_data_block)
  
  # Add return levels to the dataframe
  return_levels_all_block <- rbind(return_levels_all_block,
                                   data.frame(station_name = station_data_block$station_name %>% unique(),
                                              return_2yr = return_levels_block[1],
                                              return_5yr = return_levels_block[2],
                                              return_20yr = return_levels_block[3],
                                              return_100yr = return_levels_block[4]))
}

# Print the dataframe
print(return_levels_all_block %>% filter(station_name == "795"))  

```

```{r}
# Define the GEV parameters
location <- 0.7659
scale <- 0.1167
shape <- -0.0274



# Create a sequence of x values over which to plot the GEV distribution
x_values <- seq(-1, 2, length.out = 1000)

gev_pdf <- extRemes::devd(x_values, loc = location, scale = scale, shape = shape, type = "GEV")

# Plot the GEV distribution using base R plotting functions
plot(x_values, gev_pdf, type = "l", lwd = 2, col = "blue",
     main = "Generalized Extreme Value (GEV) Distribution",
     xlab = "Value", ylab = "Density")



```

## Map the return levels 

```{r}

#Load spatial data
ireland <- st_read("IRL_adm0.shp")


data_unique <- distinct(data, station_name, lon, lat)

## switched to use inner join to join the datasets
return_levels_with_coords_block <- inner_join(return_levels_all_block, data_unique, by = "station_name")

# Convert return_levels_all_with_coords to sf object
return_levels_sf_block <- st_as_sf(return_levels_with_coords_block, coords = c("lon", "lat"), crs = 4326)

# Plot map of Ireland with return levels represented by colors
tm_shape(ireland) +
  tm_borders() +
  tm_shape(return_levels_sf_block) +
  tm_dots(col = "return_2yr", size = 1.5, alpha = 0.5, title = "Return Level (2yr) Using Block Maxima") +
  tm_style("gray")

# Plot map of Ireland with return levels represented by colors
tm_shape(ireland) +
  tm_borders() +
  tm_shape(return_levels_sf_block) +
  tm_dots(col = "return_5yr", size = 1.5, alpha = 0.5, title = "Return Level (5yr) Using Block Maxima") +
  tm_style("gray")


# Plot map of Ireland with return levels represented by colors
tm_shape(ireland) +
  tm_borders() +
  tm_shape(return_levels_sf_block) +
  tm_dots(col = "return_20yr", size = 1.5, alpha = 0.5, title = "Return Level (20yr) Using Block Maxima") +
  tm_style("gray")

# Plot map of Ireland with return levels represented by colors
tm_shape(ireland) +
  tm_borders() +
  tm_shape(return_levels_sf_block) +
  tm_dots(col = "return_100yr", size = 1.5, alpha = 0.5, title = "Return Level (100yr) Using Block Maxima") +
  tm_style("gray")
```
```{r}

library(cluster)

# Extract the coordinates from the 'geometry' column
clustering_data <- st_coordinates(return_levels_sf_block)

# Create a data frame with return levels and extracted coordinates
clustering_data <- data.frame(
                              lon = clustering_data[, "X"],
                              lat = clustering_data[, "Y"])

# Standardize the data (optional but recommended for K-means)
scaled_data <- scale(clustering_data)

# Perform K-means clustering
k <- 4  # Number of clusters
set.seed(123)  # Set seed for reproducibility
kmeans_result <- kmeans(scaled_data, centers = k)

# Define cluster names
cluster_names <- c("West", "South", "North", "East")

# Add cluster names to the original data
return_levels_with_coords_block <- return_levels_sf_block %>%
  mutate(cluster = factor(kmeans_result$cluster, labels = cluster_names))

# Convert to sf object with clusters included
return_levels_sf_clusters <- st_as_sf(return_levels_with_coords_block, coords = c("lon", "lat"), crs = 4326)

# Custom colors for clusters
custom_colors <- c("white", "orange", "red", "yellow")

# Plot the clusters
tm_shape(ireland) +
  tm_borders() +
  tm_shape(return_levels_sf_clusters) +
  tm_dots(col = "cluster", palette = custom_colors, size = 1, alpha = 0.5, title = "Visual Coastal Groups", title.size = 0.5) + 
  tm_style("gray")

```
## Annual exceedances of 5yr return level

I've simplified this code quite alot. Values look much more reasonable now. Now try your heat map again :) 

```{r}

exceedance_data_block <-inner_join(data %>% select(station_name, year, month, day, surge_daily_max), return_levels_with_coords_block, by = "station_name")

annual_exceedance_counts <- exceedance_data_block %>% 
                              group_by(station_name, year) %>% 
                              summarise(exceedance_5yr = sum(surge_daily_max > return_5yr),
                                        exceedance_20yr = sum(surge_daily_max > return_20yr)) %>% 
                              ungroup()

## plot for DL 
ggplot(annual_exceedance_counts %>% filter(station_name == "795"), aes(x = year, y = exceedance_5yr)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Exceedances Above 5-Year Return Level Per Year",
       x = "Year", y = "Exceedances")

```


```{r}
##FOR 20-YEAR EXCEEDANCES

# Calculate total exceedances for each station
station_exceedances_block_20yr <- exceedance_data_block %>%
  group_by(cluster, station_name) %>%
  summarise(total_exceedance_20yr = sum(surge_daily_max > return_20yr, na.rm = TRUE)) %>%
  ungroup()
# Summarize exceedances by cluster
cluster_exceedance_summary_20yr <- station_exceedances_block_20yr %>%
  group_by(cluster) %>%
  summarise(total_exceedance_20yr = sum(total_exceedance_20yr),
            station_count = n_distinct(station_name),
            avg_exceedance_20yr_per_station = total_exceedance_20yr / station_count )   %>%
  ungroup()

# Print the summary
print(cluster_exceedance_summary_20yr)

##5 YEAR EXCEEDANCES

# Calculate total exceedances for each station
station_exceedances_block_5yr <- exceedance_data_block %>%
  group_by(cluster, station_name) %>%
  summarise(total_exceedance_5yr = sum(surge_daily_max > return_5yr, na.rm = TRUE)) %>%
  ungroup()
# Summarize exceedances by cluster
cluster_exceedance_summary_5yr <- station_exceedances_block_5yr  %>%
  group_by(cluster) %>%
  summarise(total_exceedance_5yr = sum(total_exceedance_5yr),
            station_count = n_distinct(station_name),
            avg_exceedance_5yr_per_station = total_exceedance_5yr / station_count )   %>%
  ungroup()

# Print the summary
print(cluster_exceedance_summary_5yr)

##FOR 2 YEAR EXCEEDANCES

# Calculate total exceedances for each station
station_exceedances_block_2yr <- exceedance_data_block %>%
  group_by(cluster, station_name) %>%
  summarise(total_exceedance_2yr = sum(surge_daily_max > return_2yr, na.rm = TRUE)) %>%
  ungroup()
# Summarize exceedances by cluster
cluster_exceedance_summary_2yr <- station_exceedances_block_2yr %>%
  group_by(cluster) %>%
  summarise(total_exceedance_2yr = sum(total_exceedance_2yr),
            station_count = n_distinct(station_name),
            avg_exceedance_2yr_per_station = total_exceedance_2yr / station_count )   %>%
  ungroup()

# Print the summary
print(cluster_exceedance_summary_2yr)
```

```{r}
library(dplyr)
library(ggplot2)


# Step 1: Summarize total exceedances above the 5-year return level for each station and year
annual_exceedance_counts <- exceedance_data_block %>%
  group_by(cluster, station_name, year) %>%
  summarise(exceedance_5yr = sum(surge_daily_max > return_5yr, na.rm = TRUE)) %>%
  ungroup()

# Check the summarized data
print(annual_exceedance_counts)

# Step 2: Order stations by cluster and station name
station_order <- annual_exceedance_counts %>%
  arrange(cluster, station_name) %>%
  pull(station_name) %>%
  unique()

# Convert station_name to factor with the specified order
annual_exceedance_counts$station_name <- factor(annual_exceedance_counts$station_name, levels = station_order)

# Step 3: Create the heatmap plot including every year label and smaller facet labels
heatmap_plot <- ggplot(annual_exceedance_counts, aes(x = year, y = station_name, fill = exceedance_5yr)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Exceedances") +
  labs(title = "Exceedances Above 5-Year Return Level Heatmap (Block Maxima)",
       x = "Year", y = "Station") +
  theme(
    axis.text.y = element_text(size = 0),  # Adjust y-axis text size
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),  # Rotate and adjust x-axis text size
    plot.title = element_text(size = 14, face = "bold"),  # Adjust title size and style
    legend.position = "bottom",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt"),  # Adjust plot margins
    strip.text.y = element_text(size = 8),  # Adjust facet label text size
    strip.background = element_rect(color = "black", fill = "lightgrey")  # Optional: Change facet background
  ) +
  scale_x_continuous(breaks = seq(min(annual_exceedance_counts$year), max(annual_exceedance_counts$year), by = 1)) +  # Label every year
  facet_grid(cluster ~ ., scales = "free_y", space = "free_y")  # Separate clusters into facets

# Print the plot
print(heatmap_plot)


##FOR 2YR HEATMAP
# Step 1: Summarize total exceedances above the 2-year return level for each station and year
annual_exceedance_counts_2yr_block <- exceedance_data_block %>%
  group_by(cluster, station_name, year) %>%
  summarise(exceedance_2yr = sum(surge_daily_max > return_2yr, na.rm = TRUE)) %>%
  ungroup()

# Check the summarized data
print(annual_exceedance_counts_2yr_block)

# Step 2: Order stations by cluster and station name
station_order <- annual_exceedance_counts_2yr_block %>%
  arrange(cluster, station_name) %>%
  pull(station_name) %>%
  unique()

# Convert station_name to factor with the specified order
annual_exceedance_counts_2yr_block$station_name <- factor(annual_exceedance_counts_2yr_block$station_name, levels = station_order)

# Step 3: Create the heatmap plot including every year label and smaller facet labels
heatmap_plot_2yr_block <- ggplot(annual_exceedance_counts_2yr_block, aes(x = year, y = station_name, fill = exceedance_2yr)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Exceedances") +
  labs(title = "Exceedances Above 2-Year Return Level Heatmap (Block Maxima)",
       x = "Year", y = "Station") +
  theme(
    axis.text.y = element_text(size = 0),  # Adjust y-axis text size
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),  # Rotate and adjust x-axis text size
    plot.title = element_text(size = 14, face = "bold"),  # Adjust title size and style
    legend.position = "bottom",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt"),  # Adjust plot margins
    strip.text.y = element_text(size = 8),  # Adjust facet label text size
    strip.background = element_rect(color = "black", fill = "lightgrey")  # Optional: Change facet background
  ) +
  scale_x_continuous(breaks = seq(min(annual_exceedance_counts_2yr_block$year), max(annual_exceedance_counts_2yr_block$year), by = 1)) +  # Label every year
  facet_grid(cluster ~ ., scales = "free_y", space = "free_y")  # Separate clusters into facets

# Print the plot
print(heatmap_plot_2yr_block)

#########################################FOR 20YR HEATMAP
# Step 1: Summarize total exceedances above the 20-year return level for each station and year
annual_exceedance_counts_20yr_block <- exceedance_data_block %>%
  group_by(cluster, station_name, year) %>%
  summarise(exceedance_20yr = sum(surge_daily_max > return_20yr, na.rm = TRUE)) %>%
  ungroup()

# Check the summarized data
print(annual_exceedance_counts_20yr_block)

# Step 2: Order stations by cluster and station name
station_order <- annual_exceedance_counts_20yr_block %>%
  arrange(cluster, station_name) %>%
  pull(station_name) %>%
  unique()

# Convert station_name to factor with the specified order
annual_exceedance_counts_20yr_block$station_name <- factor(annual_exceedance_counts_20yr_block$station_name, levels = station_order)

# Step 3: Create the heatmap plot including every year label and smaller facet labels
heatmap_plot_20yr_block <- ggplot(annual_exceedance_counts_20yr_block, aes(x = year, y = station_name, fill = exceedance_20yr)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Exceedances") +
  labs(title = "Exceedances Above 20-Year Return Level Heatmap (Block Maxima)",
       x = "Year", y = "Station") +
  theme(
    axis.text.y = element_text(size = 0),  # Adjust y-axis text size
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),  # Rotate and adjust x-axis text size
    plot.title = element_text(size = 14, face = "bold"),  # Adjust title size and style
    legend.position = "bottom",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt"),  # Adjust plot margins
    strip.text.y = element_text(size = 8),  # Adjust facet label text size
    strip.background = element_rect(color = "black", fill = "lightgrey")  # Optional: Change facet background
  ) +
  scale_x_continuous(breaks = seq(min(annual_exceedance_counts_20yr_block$year), max(annual_exceedance_counts_20yr_block$year), by = 1)) +  # Label every year
  facet_grid(cluster ~ ., scales = "free_y", space = "free_y")  # Separate clusters into facets

# Print the plot
print(heatmap_plot_20yr_block)

#########################################FOR 100YR HEATMAP
# Step 1: Summarize total exceedances above the 100-year return level for each station and year
annual_exceedance_counts_100yr_block <- exceedance_data_block %>%
  group_by(cluster, station_name, year) %>%
  summarise(exceedance_100yr = sum(surge_daily_max > return_100yr, na.rm = TRUE)) %>%
  ungroup()

# Check the summarized data
print(annual_exceedance_counts_100yr_block)

# Step 2: Order stations by cluster and station name
station_order <- annual_exceedance_counts_100yr_block %>%
  arrange(cluster, station_name) %>%
  pull(station_name) %>%
  unique()

# Convert station_name to factor with the specified order
annual_exceedance_counts_100yr_block$station_name <- factor(annual_exceedance_counts_100yr_block$station_name, levels = station_order)

# Step 3: Create the heatmap plot including every year label and smaller facet labels
heatmap_plot_100yr_block <- ggplot(annual_exceedance_counts_100yr_block, aes(x = year, y = station_name, fill = exceedance_100yr)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Exceedances") +
  labs(title = "Exceedances Above 100-Year Return Level Heatmap (Block Maxima)",
       x = "Year", y = "Station") +
  theme(
    axis.text.y = element_text(size = 0),  # Adjust y-axis text size
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),  # Rotate and adjust x-axis text size
    plot.title = element_text(size = 14, face = "bold"),  # Adjust title size and style
    legend.position = "bottom",
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt"),  # Adjust plot margins
    strip.text.y = element_text(size = 8),  # Adjust facet label text size
    strip.background = element_rect(color = "black", fill = "lightgrey")  # Optional: Change facet background
  ) +
  scale_x_continuous(breaks = seq(min(annual_exceedance_counts_100yr_block$year), max(annual_exceedance_counts_100yr_block$year), by = 1)) +  # Label every year
  facet_grid(cluster ~ ., scales = "free_y", space = "free_y")  # Separate clusters into facets

# Print the plot
print(heatmap_plot_100yr_block)

```

